---
title: "CodeBook"
output: html_document
---
Getting and Cleaning Data 
===============
**Human Activity Recognition Using Smartphones Data Set**

1. Merging the training and the test sets to create one data set.
The first step is to get the data. The following code creates a directory called "data" if it 
        does not already exist, then downloads, stores and unzips the UCI HAR Dataset.
```{r}
if (dir.exists("./data") == FALSE) {
         dir.create("./data")
         }
fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(fileUrl, destfile = "./data/dataset.zip")
 
unzip("./data/dataset.zip", exdir = "./data")
```
Basically, there are three major parts that need to be assembled into a single piece.
        - The first one consists of `X_train.txt` and `X_test.txt`
                We read the files into R using `read.table()`
```{r} 
                dataXtrain <- read.table("./data/UCI HAR Dataset/train/X_train.txt", header = FALSE) 
                dataXtest <- read.table("./data/UCI HAR Dataset/test/X_test.txt", header = FALSE)
```
                We combine the two using `rbind()`
```{r}
                combined_X <- rbind(dataXtrain, dataXtest)
```
                and name the variables using `features.txt`
```{r}
                features_names <- read.table("./data/UCI HAR Dataset/features.txt" , sep = " ", header = FALSE)
                names(combined_X) <- features_names$V2
```
        - The second part is comprised of `Y_train.txt` and `Y_test.txt`. We name this column `activity`
```{r}
                dataYtrain <- read.table("./data/UCI HAR Dataset/train/Y_train.txt" , header = FALSE)
                dataYtest <- read.table("./data/UCI HAR Dataset/test/Y_test.txt" , header = FALSE)
                combined_Y <- rbind(dataYtrain, dataYtest)
                names(combined_Y) <- c("activity")
```
        - The third part is a `Subject` column. Again, we repeat the same procedure and label the variable as "Subject"
```{r}
                subject_train <- read.table("./data/UCI HAR Dataset/train/subject_train.txt" , header = FALSE)
                subject_test <- read.table("./data/UCI HAR Dataset/test/subject_test.txt" , header = FALSE)
                combined_subject <- rbind(subject_train, subject_test)
                names(combined_subject) <- c("Subject")
```
Now it's time to join three pieces together using `cbind()`
```{r}
combined_XY <- cbind(combined_X, combined_Y) 

mydata <- cbind(combined_XY, combined_subject)
```
Check the contents of `mydata`
```{r}
str(mydata)
```
Remove the objects we no longer need
```{r}
rm("dataXtrain", "dataXtest", "dataYtrain", "dataYtest", "subject_train", "subject_test")
rm("combined_XY", "combined_X", "combined_Y", "combined_subject")
```
2. Extracting only the measurements on the mean and standard deviation for each measurement.  
Using `grep()` function, create two vectors containing only `mean()` and `std()` character strings in `features_names`. In this case, I decided to include `freqMean` and `angle` data as well.
```{r}
contains_mean <- grep("mean()", features_names$V2, ignore.case = TRUE, value = TRUE)
contains_sd <- grep("std()", features_names$V2, ignore.case = TRUE, value = TRUE)
```
Create a vector of variable names containing only `mean` and `std` values +"Subject" and "activity". `mean_sd_names_vector` will be used for subsetting `mydata` 
```{r}
mean_sd_names_vector <- c(contains_mean, contains_sd, "Subject", "activity")
```
Print out the vector
```{r}
mean_sd_names_vector
```
Subset `mydata` using `mean_sd_names_vector`
```{r}
mydata_mean_sd <- subset(mydata, select = mean_sd_names_vector)
```
Check the contents
```{r}
str(mydata_mean_sd)
```
3. Adding descriptive activity names to name the numbered activities in the data set.
Merging data by `Activity`
First we need to load `activity_labels.txt` file into R as `data.frame`
```{r}
activities_names <- read.table("./data/UCI HAR Dataset/activity_labels.txt", sep=" ", header =FALSE)
activities_names <- as.data.frame(activities_names)
```
We name the two columns as "id" and "Activity"
```{r}
names(activities_names) <- c("id", "Activity")
activities_names
```
Now we are ready to `merge()` the two by `activity` and `id`
```{r}
mergedData <- merge(mydata_mean_sd, activities_names, by.x = "activity", by.y = "id")
str(mergedData)

```
Pull up `dplyr` package and remove the duplicate column `activity` 
```{r}
library(dplyr)
Data <- select(mergedData, -(activity))
```
Rearrange the order of `Subject` and `Activity` columns. Check the data.
```{r}
Data <- Data[c(87, 88, 1:86)]
str(Data)
```
4. Appropriately labeling the data set with descriptive variable names.
Change the variable names to make them easier to understand.
Use function `gsub()`. Reference: http://www.cookbook-r.com/Manipulating_data/Renaming_columns_in_a_data_frame/
Find variable descriptions in `features_info.txt`. Change "Gyro" to "Gyroscope", "Acc" to "Accelerometer", "tBody" to "TimeBody", "tGravity" to "TimeGravity".
```{r}
names(Data) <- gsub("Gyro", "Gyroscope", names(Data))
names(Data) <- gsub("Acc", "Accelerator", names(Data))
names(Data) <- gsub("tBody", "TimeBody", names(Data))
names(Data) <- gsub("tGravity", "TimeGravity", names(Data))
```

Change "Freq" to "Frequency", "fBody" to "FrequencyBody", "Mag" to "Magnitude", "Jerk" to "Jerk-signal", "BodyBody" to "Body".
```{r}
names(Data) <- gsub("Freq", "Frequency", names(Data))
names(Data) <- gsub("fBody", "FrequencyBody", names(Data))
names(Data) <- gsub("Mag", "Magnitude", names(Data))
names(Data) <- gsub("Jerk", "Jerk-signal", names(Data))
names(Data) <- gsub("BodyBody", "Body", names(Data))
```
Rename the axes.
```{r}
names(Data) <- gsub("-X", "-X-axis", names(Data))
names(Data) <- gsub("-Y", "-Y-axis", names(Data))
names(Data) <- gsub("-Z", "-Z-axis", names(Data))
```
Check the names
```{r}
names(Data)
```
5. Our last step is to produce a tidy dataset with the average of each variable for each activity and each subject.
Using `group_by` function from `dplyr` we group `Data` by `Subject` and `Activity`.
```{r}
groupedData <- group_by(Data, Subject, Activity)
```
Using `summarise_each` (Reference: http://stackoverflow.com/questions/21644848/summarizing-multiple-columns-with-dplyr) we apply the `mean()` function to each column of the `Data`.
```{r}
tidyData <- summarise_each(groupedData, funs(mean))
```
Check `tidyData`
```{r}
str(tidyData)
```
Then we write the table to a txt file
```{r}
write.table(tidyData, file = "tidyData.txt", row.name=FALSE, col.names = TRUE)
```
To load the 'tidyData.txt' use 
```{r}
tidyData <- read.table("tidyData.txt", header = TRUE)
```